---
title: "Health Care Analytics"
author: "Vamsi Tallapu Reddy, Sujan Barama, Atharva Dastane, Anuradha Mysore Ravishankar, Anita T"
date: "9/23/2021"
output: pdf_document
---

# Final Project DTSC 5301-001: Data Science as a Field

# Key Question - How have global health trends changed over the last 2 decades?
## What are the basic health metric trends?
## What are the major disease trends?
## How have vaccination rates changed over time?
## What is the financial impact caused by these changes?

### Loading all the data

Data Source:https://apps.who.int//nha//database//Home//IndicatorsDownload//en

Other data sources:

1. https://www.who.int/teams/immunization-vaccines-and-biologicals/vaccine-access/planning-and-financing/immunization-financing-indicators

2. https://www.kaggle.com/utkarshxy/who-worldhealth-statistics-2020-complete

github link - https://github.com/adastane100/DTSC5301_PROJECT.git
---
### Short data description

WHO data - data of health expenditure of 190 countries at a country x year level. Sample columns are; GDP, population, current health expenditure as a % of GDP, out of pocket expenses, voluntary health insurance, etc.

Kaggle data - No of doctors, nurses, life expectancy, mortality rate, tuberculosis, malaria cases, etc at a country year level

WHO immunization data - Amount spent on immunization at a country year level
---
### Code starts here:

Loading libraries

```{r, echo = FALSE}

library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggpubr)
library(readxl)
```
---


### Load the main WHO data & clean data


```{r,echo=FALSE}

who_url <- "https://apps.who.int//nha//database//Home//IndicatorsDownload//en"
who_download<- download.file(url = who_url, destfile = "data_Who.xlsx", mode="wb")
master_data <- read_excel("data_Who.xlsx")
exp<-"https://cdn.who.int/media/docs/default-source/immunization/financing/3-total-expenditure-routine-immunization.xlsx?sfvrsn=517de01c_4"
exp_download<- download.file(url = exp, destfile = "exp_who.xlsx", mode="wb")
immu_data <- read_excel("exp_who.xlsx", skip=2)

#Keep the relevant data. Filtering the data
newdata <- master_data %>% select(country=country,
                                  year=year,
                                  income_group=`income group (2018)`,
                                  region = `region (WHO)`,phc_che = phc_che,
                                  phc_usd_pc = phc_usd_pc,
                                  gdp = gdp,
                                  pop = pop,
                                  che_gdp = che_gdp,
                                  che_pc_usd = che_pc_usd,
                                  vhi_che = vhi_che,
                                  oops_che = oops_che)
newdata
```
---

### Create functions for everything:


```{r}
#To load data
load_data <-function(x){
  data<- read.csv(x)
}
#To select relevant data
filter_data <- function(df){
  if ("Dim1" %in% colnames(df) == TRUE){
    data <- filter(df, Dim1 == "Both sexes")
    data <- select(data, country = Location, year = Period, value = "First Tooltip")
  }
  else{
    data <- select(df, country = Location, year = Period, value = "First Tooltip")
  }
}

```

---

### Extracting the data from the URL 


```{r}
url_in <-"https://raw.githubusercontent.com/adastane100/DTSC5301_PROJECT/main/Data/"
file_names <- c("crudeSuicideRates.csv","infantMortalityRate.csv","lifeExpectancyAtBirth.csv",
                "medicalDoctors.csv","neonatalMortalityRate.csv","nursingAndMidwife.csv",
                "pharmacists.csv","under5MortalityRate.csv","30-70cancerChdEtc.csv","newHivInfections.csv","incedenceOfTuberculosis.csv","incedenceOfMalaria.csv","hepatitusBsurfaceAntigen.csv",
                "3-total-expenditure-routine-immunization.xlsx")
urls <- str_c(url_in,file_names)
```
---

### Loading Kaggle data from the github

```{r}
suicides <- read_csv(urls[1], show_col_types = FALSE)
mortality <- read_csv(urls[2], show_col_types = FALSE)
life_expectancy <- read_csv(urls[3], show_col_types = FALSE)
doctors <- read_csv(urls[4], show_col_types = FALSE)
neo_mortality <- read_csv(urls[5], show_col_types = FALSE)
nursing <- read_csv(urls[6], show_col_types = FALSE)
pharmacists <- read_csv(urls[7], show_col_types = FALSE)
kids_mortality <- read_csv(urls[8], show_col_types = FALSE)
cancer_cases <- read_csv(urls[9], show_col_types = FALSE)
hiv_cases <- read_csv(urls[10], show_col_types = FALSE)
tuberculosis_cases <- read_csv(urls[11], show_col_types = FALSE)
malaria_cases <- read_csv(urls[12], show_col_types = FALSE)
hepatitis_cases <- read_csv(urls[13], show_col_types = FALSE)

```
---


### Transform the data that is extracted from Github and rename the columns in the main data set


```{r}
#Transform Data
mortality_clean <- filter_data(mortality)
life_expectancy_clean <- filter_data(life_expectancy)
doctors_clean <- filter_data(doctors)
neo_mortality_clean <- filter_data(neo_mortality)
nursing_clean <- filter_data(nursing)
pharmacists_clean <- filter_data(pharmacists)
kids_mortality_clean <- filter_data(kids_mortality)
cancer_cases_clean <- filter_data(cancer_cases)
hiv_cases_clean <- filter_data(hiv_cases)
tuberculosis_cases_clean <- filter_data(tuberculosis_cases)
malaria_cases_clean <- filter_data(malaria_cases)
hepatitis_cases_clean <- filter_data(hepatitis_cases)

# Rename value column

mortality_clean <- rename(mortality_clean, mortality_rate = value)
life_expectancy_clean <- rename(life_expectancy_clean, life_expectancy = value)
doctors_clean <- rename(doctors_clean, doctors = value)
neo_mortality_clean <- rename(neo_mortality_clean, neo_mortality_rate = value)
nursing_clean <- rename(nursing_clean, nurses_midwives = value)
pharmacists_clean <- rename(pharmacists_clean, pharmacists = value)
kids_mortality_clean <- rename(kids_mortality_clean, suicide_rate = value)
nursing_clean <- filter(nursing_clean, year >= 2000)
cancer_cases_clean <- rename(cancer_cases_clean, cancer_cases = value)
hiv_cases_clean <- rename(hiv_cases_clean, hiv_cases = value)
tuberculosis_cases_clean <- rename(tuberculosis_cases_clean, tuberculosis_cases = value)
malaria_cases_clean <- rename(malaria_cases_clean, malaria_cases = value)
hepatitis_cases_clean <- rename(hepatitis_cases_clean, hepatitis_cases = value)
```
---

### Merge all dataframes to create a master dataframe


```{r}
merge_cols = c("country, year")
merge1 <- merge(newdata, kids_mortality_clean, on = merge_cols, all.x = TRUE)
nrow(newdata)
merge2 <- merge(merge1, pharmacists_clean, on = merge_cols, all.x = TRUE)
merge23 <- merge(merge2, nursing_clean, on = merge_cols, all.x = TRUE)
merge3 <- merge(merge23, neo_mortality_clean, on = merge_cols, all.x = TRUE)
new_merge <- merge(merge3, doctors_clean, on = merge_cols, all.x = TRUE)
mer1 <- merge(new_merge, cancer_cases_clean, on = merge_cols, all.x = TRUE)
mer2 <- merge(mer1, tuberculosis_cases_clean, on = merge_cols, all.x = TRUE)
mer3 <- merge(mer2, malaria_cases_clean, on = merge_cols, all.x = TRUE)
mer4 <- merge(mer3, hiv_cases_clean, on = merge_cols, all.x = TRUE)
mer5 <- merge(mer4, hepatitis_cases_clean, on = merge_cols, all.x = TRUE)
new_merge2 <- merge(mer5, life_expectancy_clean, on = merge_cols, all.x = TRUE)
master_df <- merge(new_merge2, mortality_clean, on = merge_cols, all.x = TRUE)
#master_df <- merge(new_merge3, suicides_clean, on = merge_cols, all.x = TRUE)

```
---

### Transform data further to begin analysis

```{r}

master_df["health_expenditure"] = (master_df$che_gdp/100)*master_df$gdp 
master_df <- filter(master_df, income_group == "Low" | income_group == "Low-Mid")
master_df$che <- substr(master_df$suicide_rate,1,5)
master_df$mortality_rate <- substr(master_df$mortality_rate,1,5)
master_df$suicide_rate <- as.numeric(master_df$suicide_rate)
master_df$mortality_rate <- as.numeric(master_df$mortality_rate)
```
---

### How have doctor numbers changed from 2000 to 2018?

```{r}

# Filtering data for 2000 & 2018 to compare the two
doc2000 <- filter(master_df, year == 2000 | year == 2001, doctors >= 0)
doc2000 <- select(doc2000, country, year, doctors)

doc2018 <- filter(master_df, year == 2017 | year == 2018, doctors >= 0)
doc2018 <- select(doc2018, country, year, doctors)

#Selecting countries that are common for 2000,2001 and 2017,2018 as comparison can only be made when the countries are similar

countries_2018 <- unique(doc2018$country)
countries_2000 <- unique(doc2000$country)

unique_countries = countries_2000[countries_2000 %in% countries_2018]

#Checking mean of doctors for 2000 for common countries
doc2000 <- filter(doc2000, country == "Bangladesh" | country == "Chad" | country == "Honduras" |
         country == "India" | country == "Lao People's Democratic Republic" | country == "Pakistan" |
         country == "Papua New Guinea" |
         country == "Tunisia" | country == "Zimbabwe")

mean(doc2000$doctors)


#Checking mean of doctors for 2018 for common countries
doc_2018 <- filter(doc2018, country == "Bangladesh" | country == "Chad" | country == "Honduras" |
                     country == "India" | country == "Lao People's Democratic Republic" | country == "Pakistan" |
                     country == "Papua New Guinea"  | 
                     country == "Tunisia" | country == "Zimbabwe")

doc_2018
mean(doc_2018$doctors)
 #5.5

# Biggest Winners - Bangladesh, India
```
---

# How have pharmacists changed from 2000 to 2018?

```{r}

# Filtering data for 2000 & 2018 to compare the two
pharm2000 <- filter(master_df, year == 2000| year == 2001 | year == 2002,  nurses_midwives >= 0)
pharm2000 <- select(pharm2000, country, year, pharmacists_2000 = pharmacists)

pharm2018 <- filter(master_df, year == 2016 | year == 2017 | year == 2018, nurses_midwives >= 0)
pharm2018 <- select(pharm2018, country, year, pharmacists_2018 = pharmacists)

#checking for common countries among the two sets
countries_2018 <- unique(pharm2018$country)
countries_2000 <- unique(pharm2000$country)

head(countries_2018)

unique_countries = countries_2000[countries_2000 %in% countries_2018]
head(unique_countries)

pharmacy <- merge(pharm2000, pharm2018, all = TRUE)
head(pharmacy)

#Biggest Winners - Nepal, East Timor 

```
---

### How have nurses/midwives changed from 2000 and 2018

```{r}

#Filtering out data for 2000 and 2018
nurses2000 <- filter(master_df, year == 2002| year == 2003 | year == 2004,  pharmacists >= 0)
nurses2000 <- select(nurses2000, country, year, nurses_2000 = nurses_midwives)

nurses2018 <- filter(master_df, year == 2016 |year == 2017 | year == 2018, pharmacists >= 0)
nurses2018 <- select(nurses2018, country, year, nurses_2018 = nurses_midwives)

# Performing an inner join to get common comparable data
nurses <- merge(nurses2000, nurses2018, all = TRUE)
head(nurses)
#From this table
#Biggest Winners - Nepal, Indonesia
#Change from 9.625 -  13.7
```
---

### How has health expenditure per capita changed from 2000 to 2018

```{r}
#Filtering out data for 2000 and 2018
expenditure2000 <- filter(master_df, year == 2002| year == 2003 | year == 2004,  che_pc_usd >= 0)
expenditure2000 <- select(expenditure2000, country, year, health_expenditure_2000 = che_pc_usd)

expenditure2018 <- filter(master_df, year == 2016 |year == 2017 | year == 2018, che_pc_usd >= 0)
expenditure2018 <- select(expenditure2018, country, year, health_expenditure_2018 = che_pc_usd)

#Inner join to get common countries
expenditure <- merge(expenditure2000, expenditure2018, all = TRUE)
head(expenditure)
#From the table
#Biggest Winners - Sudan, Myanmar ~300 dollars/person, No point in this

#Change from 32.78 - 100
```
---

# How has life expectancy changed from 2000 to 2018

```{r}

#Filter out data for 2000 and 2018
life_expectancy2000 <- filter(master_df, year == 2000| year == 2005 ,  life_expectancy >= 0)
life_expectancy2000 <- select(life_expectancy2000, country, year, life_expectancy_2000 = life_expectancy)

life_expectancy2018 <- filter(master_df, year == 2015 |year == 2019 , life_expectancy >= 0)
life_expectancy2018 <- select(life_expectancy2018, country, year, life_expectancy2018 = life_expectancy)

#Inner join to get common countries
expectancy <- merge(life_expectancy2000, life_expectancy2018, all = TRUE)
head(expectancy)
#latest data available was 2015
#From the table
#Biggest Winners - Tunisia & Bangladesh, Rwanda & Burundi  - Tunisia ~75, Malaysia, South Africa 
#Change from 58 - 65
```
---

### How has the trends in health expenditure impacted Tuberculosis?

```{r}
# Calculation of Current Health Expenditure from GDP
master_df<-master_df%>% mutate(CurrentHealthExp= (che_gdp*gdp)/100)

#Preparing the Required Data
gfg <- master_df$tuberculosis_cases
res = as.numeric(gsub(".*?([0-9]+).*", "\\1", gfg))
tuberculosis<-master_df%>%
  mutate(tuberculosis_cases_per_1000_pop = res)%>%
  group_by(year)%>%
  summarize(m=mean(tuberculosis_cases_per_1000_pop,na.rm=TRUE))
che<- master_df %>% group_by(year)%>%
  summarise(che1=log(mean(CurrentHealthExp,na.rm=TRUE)))

f<-merge(tuberculosis,che,on=year)
```

```{r}
#Plotting health expenditure vs Tuberculosis per 1000 thousand population
g2<-ggplot(data = f, aes(x = year,group=1)) +
    geom_line(aes(y= m,group=1),color= "blue")+
  geom_line(aes(y=che1,group=1), color="green")

```


```{r}
#Analyzing and Plotting Region Wise from 2000 to 2018 for TB per thousand vs Health Expenditure
master_df<-master_df%>%
  mutate(tuberculosis_cases_per_1000_pop = res)
region<- master_df%>%filter(year==2000)%>% group_by(region)%>% summarize (tb= mean(tuberculosis_cases_per_1000_pop, na.rm=TRUE))
region1<- master_df%>%filter(year==2000)%>% group_by(region)%>% summarize (che= mean(CurrentHealthExp, na.rm=TRUE))
ggp <- ggplot(region, aes(x = region, y = tb,group=1)) +  # Create stacked bar chart
  geom_bar(stat = "identity",  fill= 'Blue')
ggp
ggp1<- ggplot(region1,aes(x = region, y=che,group=1))+
  geom_bar(stat="identity", fill='Green')
ggp 
ggp1

                     
```

```{r}

region2018<- master_df%>%filter(year==2018)%>% group_by(region)%>% summarize (tb= mean(mean(tuberculosis_cases_per_1000_pop, na.rm=TRUE)))
region2<- master_df%>%filter(year==2018)%>% group_by(region)%>% summarize (che= mean(CurrentHealthExp, na.rm=TRUE))
ggp2018 <- ggplot(region2018, aes(x = region, y = tb,group=1) ) +  # Create stacked bar chart
  geom_bar(stat = "identity", fill='Blue')
ggp2018
ggpche<- ggplot(region2,aes(x = region, y=che,group=1))+
  geom_bar(stat="identity", fill= 'Green')
ggpche
```
---
```{r}
library('ggpubr')
figure= ggarrange(ggp1,ggpche,ggp,ggp2018,labels=c('a','b','c','d'), ncol=2, nrow=2)

```

Conclusion:
As the current health expenditure increased from 2000 to 2018, the tuberculosis cases decreased from 2000 to 2018 but we can see that the African and SEAR regions didn't show much improvement because in Africa, the patients who died due to TB has HIV and are considered as TB patients instead OF HIV. In SEAR, there is a country named Combodia in which 64% of people suffer from TB which increases the overall cases in TB. The bias in this is 40% of people are tested negative though they have the TB which makes the TB to spread. 

---

### How have trends in Malaria changed from 2000 to 2018 with the growing Heath Expenditure?

```{r}
#Calculating the Actual Current Health Expenditure From GDP
master_df<-master_df%>%mutate(CurrentHealthExp = (che_gdp/100)*gdp)

#Plotting for Current Health Expenditure vs Malaria cases per thousand Population
malaria <- master_df %>% group_by(year)%>%
  summarise(m=mean(malaria_cases,na.rm=TRUE))
gdp1<- master_df %>% group_by(year)%>%
  summarise(gdp=log(mean(CurrentHealthExp,na.rm=TRUE)))
f<-merge(malaria,gdp1,on=year)
f
write.csv(f,'analysis.csv')
g=ggplot(data=f,aes(x=year,group=1))+geom_line(aes(y=m,group=1),color='red')+geom_line(aes(y=gdp,group=1),color='blue')
g

#Analyzing and Plotting Region Wise from 2008 to 2018 for Malaria per thousand vs Health Expenditure

region_CHE_2008 <- master_df %>%
  filter(year == 2008)%>%
  group_by(region)%>%
  summarize(m=mean(CurrentHealthExp,na.rm=TRUE))
write.csv(region_CHE_2008,'analysis1.csv')
g_che_2008=ggplot(data=region_CHE_2008,aes(x=region,y=m,group=1))+geom_bar(stat='identity',fill='blue')
g_che_2008
region_malaria_2008 <- master_df %>%
  filter(year == 2008)%>%
  group_by(region)%>%
  summarize(m=mean(malaria_cases,na.rm=TRUE))
write.csv(region_malaria_2008,'analysis2.csv')
g_mal_2008=ggplot(data=region_malaria_2008,aes(x=region,y=m,group=1))+geom_bar(stat='identity',fill='red')
g_mal_2008
region_CHE_2018 <- master_df %>%
  filter(year == 2018)%>%
  group_by(region)%>%
  summarize(m=mean(CurrentHealthExp,na.rm=TRUE))
#write.csv(region_CHE_2018,'analysis3.csv')
g_che_2018=ggplot(data=region_CHE_2018,aes(x=region,y=m,group=1))+geom_bar(stat='identity',fill='blue')
g_che_2018
region_malaria_2018 <- master_df %>%
  filter(year == 2018)%>%
  group_by(region)%>%
  summarize(m=mean(malaria_cases,na.rm=TRUE))
write.csv(region_malaria_2018,'analysis4.csv')
g_mal_2018=ggplot(data=region_malaria_2018,aes(x=region,y=m,group=1))+geom_bar(stat='identity',fill='red')
g_mal_2018
figure<-ggarrange(g_che_2008,g_che_2018,g_mal_2008,g_mal_2018,
                  ncol = 2, nrow = 2)
figure
```
---

4 important conclusions:-

1- Significant increase in health expenditure led to significant decrease in malaria cases.This is evident from the SEA and WPR regions.
2-Poor investment in health expenditure led to significant increase in malaria cases.African countries have highest number of cases than any other region.
3-Significant increase in health expenditure did not affect the malaria cases.This is evident from American region.Maybe malaria was not a disease of concern for them.
4-No data for Europe meant European regions were Malaria free.On Googling it ,we found that Europe was malaria free in 2015.

### How have vaccination rates changed over time?

```{r}

#Preparing data and filtering out Low and Low-Mid Countries.


im <- immu_data%>%filter(income == 'Low' |income == 'Low-Mid')
avgs_of_years = colMeans(im[5:18],na.rm = TRUE)
avgs_of_years_2016 = colMeans(im[5:15],na.rm = TRUE)
avgs_of_years = c(avgs_of_years)
years = c(2006,2007,2008,2009,2010,2011,2012,2013,2014,2015,2016,2017,2018,2019)
plot_of_avg <- plot(years,avgs_of_years, type = 'l', col="green", lwd=5)
typeof(avgs_of_years)
years_2016 = c(2006,2007,2008,2009,2010,2011,2012,2013,2014,2015,2016)
plot_of_avg <- plot(years_2016,avgs_of_years_2016, type = 'l',xlab = "Years",ylab = "Average of expenditure", col="green", lwd=5)
avgs_of_years_2016 = c(avgs_of_years_2016)
typeof(avgs_of_years_2016)

#BOXPLOT FOR 2017
plot_2017 <- ggplot(data = im)+geom_boxplot(aes(x=`2017`))
plot_2017

#BOXPLOT FOR 2018
plot_2018 <- ggplot(data = im)+geom_boxplot(aes(x=`2018`))
plot_2018

# LINE PLOT FOR CLOSER LOOK
avgs_of_3years = colMeans(im[15:18],na.rm = TRUE)
avgs_of_3years = c(avgs_of_3years)
typeof(avgs_of_3years)
last_years = c(2016,2017,2018,2019)
plot_of_avg <- plot(last_years,avgs_of_3years, type = 'l', col="red", lwd=5)
plot_of_avg

#WHAT HAPPEND from 2016 to 2018 , region wise divison
region_2016<-im%>%
  group_by(Region)%>%
  summarize(mean(`2016`,na.rm = TRUE))
region_2016
region_2017<-im%>%
  group_by(Region)%>%
  summarize(mean(`2017`,na.rm = TRUE))
region_2017
region_2018<-im%>%
  group_by(Region)%>%
  summarize(mean(`2018`,na.rm = TRUE))
region_2018

#From the values we can clearly see that there's a drastic drop in AFRO region and  SEARO region has increased 1/3 times but it  has sudden drop from 2017 to 2018
#Health exp trend in just AFR region.
master_df$expen = master_df$che_gdp*master_df$gdp
region_afro_exp <- master_df%>%
  filter(region == 'AFR')%>%
  group_by(year)%>%
  summarize(mean(expen,na.rm = TRUE))
region_afro_exp
plot(region_afro_exp$year,region_afro_exp$expen, type = 'l', lwd = 5)

#Top five countries in expenditure
im_678 <- im%>%
  select(`2006`,`2007`,`2008`)
im$min <- apply(im_678,1,FUN=min)
im_1516 <- im%>%
  select(`2015`,`2016`)
im$max <- apply(im_678,1,FUN=max)
im$per_change <- (im$max - im$min)/im$min
im$per_change
 im <- im[order(im$per_change, decreasing = TRUE), ]
im
# so the top five countries are per change
im%>%
  select("Country",'Region', "per_change")
#so the top five countries are in 2016
im <- im[order(im$`2016`, decreasing = TRUE), ]
im
im%>%
  select("Country",'Region',`2016`)

im <- immu_data%>%
  filter(income == 'Low' | income == 'Low-Mid')
avgs_of_years = colMeans(im[5:18],na.rm = TRUE)
avgs_of_years_2016 = colMeans(im[5:15],na.rm = TRUE)
avgs_of_years = c(avgs_of_years)
write.csv(avgs_of_years,'analysis_till_2019.csv')
typeof(avgs_of_years)
years = c(2006,2007,2008,2009,2010,2011,2012,2013,2014,2015,2016,2017,2018,2019)
plot_of_avg <- plot(years,avgs_of_years, type = 'l', col="green", lwd=5)
typeof(avgs_of_years)
years_2016 = c(2006,2007,2008,2009,2010,2011,2012,2013,2014,2015,2016)
plot_of_avg <- plot(years_2016,avgs_of_years_2016, type = 'l',xlab = "Years",ylab = "Average of expenditure", col="green", lwd=5)
```

Health Expenditure on Vaccination rates increased by 141.6% from 2006 to 2016.


---

### What is the financial impact?

Let us consider the financial impact of the WHO expenditure on the society of low mid and low income groups
The first plot clearly shows the upward trend in Voluntary health expenses and downward trend in the Out of Pocket Expenditure

```{r}
library(tidyverse)
library(readxl)
library("ggplot2")
theme_set(theme_bw())
library("sf")


#Load the main WHO data
who_url = "https://apps.who.int//nha//database//Home//IndicatorsDownload//en"
who_download<- download.file(url = who_url, destfile = "data_Who.xlsx", mode="wb")
master_data <- read_excel("data_Who.xlsx")
#Keep the relevant data

head(newdata)
# Financial impact of such schemes on individuals
master_df <- filter(newdata, income_group == "Low" | income_group == "Low-Mid")

# trends and observations in the out of pocket expenses by an individual


master_df_region <- master_df %>%
  group_by(region) %>%
  summarise(mean_vhi = mean(vhi_che, na.omit = TRUE),
            mean_oops = mean(oops_che, na.omit = TRUE))

#View(master_df1)
# Low income and low mid - Europe
# Clearly individuals in European countries have more out of pocket expenses, and so they are not making use of the voluntary health insurance
# they are not only offered to the individuals under govt jobs but also are additional benefits other than health expenses provided by the govt

master_df_country <- master_df %>%
                  group_by(country) %>%
                  summarise(mean_vhi = mean(vhi_che, na.omit = TRUE),
                  mean_oops = mean(oops_che, na.omit = TRUE))

# it is visible also at the country level

# Visuals to cross verify the numbers

voluntary_expense_trend <- ggplot(data = master_df, aes(x=year, y=vhi_che))+
  geom_bar(stat="identity",(aes(fill=region))) +
  ggtitle("Year over Year trend of the Voluntary Health Expenditure") +
  xlab("year") +
  ylab("VHI in % as Current Expenditure")
voluntary_expense_trend

out_of_expense_trend <- ggplot(data = master_df, aes(x=year, y=oops_che))+
  geom_bar(stat="identity",(aes(fill=region))) +
  ggtitle("Year over Year trend of the Out-of-Pocket Expenditure") +
  xlab("year") +
  ylab("OOPS in % as Current Expenditure")
out_of_expense_trend

 


```
This is an immense improvement in the healthcare investment by the government and less out of pocket expenses for an individual
---

### Biases in our analysis

We can see 2 different kinds of biases: biases from data sources & collection and human biases
Biases from data sources & collection
1.Data unavailability - This data consists of a lot of null values, missing values and NA values. Data for some of the fiscal years are missing, so we had to break the dataset in order to showcase the continuous trends.

2. Inaccuracies in data collection – Any and every inaccurate data point could result in erroneous results.

3. Restricted time frame – We have chosen 2000 to 2018 and this could possibly be a source of bias

Human biases –

1. Confirmation bias – bias induced by existing inherent beliefs

2.Dunning Kruger effect –  This bias leads people to view an idea or event as simplistic because they don’t have a lot of information on the subject. While under the influence of Dunning-Kruger, people overestimate their knowledge of something and it prevents them from being curious and seeking out information

3.Cultural bias – Bias introduced by cultural differences. For example, hygiene isn’t important to country X. Initially, there were assumptions regarding the less focus on AFR countries. From the inference, we see that they have been doing well in healthcare sector

## Conclusion

We would like to conclude our analysis with a relevant quote -

"_People often call me an optimist, because I show them the enormous progress they didn't know about. That makes me angry. I'm not an optimist. That makes me sound naive. I'm a very serious “possibility”. That’s something I made up. It means someone who neither hopes without reason, nor fears without reason, someone who constantly resists the over dramatic worldview. As a possibility, I see all this progress, and it fills me with conviction and hope that further progress is possible. This is not optimistic. It is having a clear and reasonable idea about how things are. It is having a worldview that is constructive and useful_"

Our analysis shows that things are genuinely getting better in global health. There are  a lot of challenges facing low and low-mid countries today, but this analysis is an acknowledgement of their efforts in improving their health in general. Some of these are -

1. There has been a 200% increment in health expenditure 
2. Life expectancy has increased to 65 from 58
3. Healthcare has become more affordable as compared to 2000 
4. Money spent on immunizations has increased by 141.6%
5. Europe has eliminated Malaria
6. Significant reduction in Tuberculosis & Malaria cases 
